Downstream analyses of whole-genome bisulfite-sequencing data

### 2015-05-04 Rmarkdown/appendix.Rmd
Fix table formatting

\begin{table}[h]
\centering
\begin{tabular}{@{}lllll@{}}
\toprule
Machine name & Processors & Cores per processor & Total number of cores & RAM    \\ \midrule
unix88       & $4\times$ Intel Xeon X7350 @ 2.93GHz & 4                   & 16                    & 128 GB \\
unix301      & $4\times$ AMD Opteron 8435 @ 2.6GHz  & 6                   & 24                    & 256 GB \\
unix302      & $4\times$ AMD Opteron 6174 @ 2.2GHz  & 12                  & 48                    & 512 GB \\
unix303      & $4\times$ AMD Opteron 6176 @ 2.3GHz  & 12                  & 48                    & 512 GB \\
unix305      & $4\times$ AMD Opteron 6276 @ 2.3GHz  & 16                  & 64                    &        \\ \bottomrule
\end{tabular}
\caption{Bioinformatics Division server specifications.}
\label{tab:unices}
\end{table}

>>>>

\begin{table}[h]
\centering
\caption{Bioinformatics Division server specifications.}
\label{tab:unices}
\begin{tabulary}{\textwidth}{LLLL}
\toprule
Machine name & Processors & Number of cores & RAM    \\ \midrule
unix88       & $4\times$ Intel Xeon X7350 @ 2.93GHz & 16 & 128 GB \\
unix301      & $4\times$ AMD Opteron 8435 @ 2.6GHz  & 24 & 256 GB \\
unix302      & $4\times$ AMD Opteron 6174 @ 2.2GHz  & 48 & 512 GB \\
unix303      & $4\times$ AMD Opteron 6176 @ 2.3GHz  & 48 & 512 GB \\
unix305      & $4\times$ AMD Opteron 6276 @ 2.3GHz  & 64 & 512 GB \\ \bottomrule
\end{tabulary}
\end{table}

### 2015-05-04 Rmarkdown/datasets.Rmd
Fix table formatting

The methylC-seq libraries from the \cite{Lister:2009hy} paper were the first published whole-genome bisulfite-sequencing libraries of mammalian DNA. A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (_IMR90_), with those from a pluripotent tissue, embryonic stem cells (_H1_). Each tissue was run in duplicate. While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described as technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing. In any case, the published analyses pool these duplicates, which ignores all between-replicate variability.

The methylC-seq libraries from the \cite{Lister:2011kg} include some created by the authors and some published by other groups (see Table \ref{tab:Lister_dataset}). These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells. There are no replicates for any of the \cite{Lister:2011kg} samples.

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset]{Samples in \emph{Lister} dataset. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{PE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
\label{tab:Lister_dataset}
\begin{tabulary}{\textwidth}{LLLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage & First published             \\ \midrule
ADS                 & Adipose                 & PE         & $75$ bp     & $23\times$    & \citet{Lister:2011kg}       \\
ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp     & $24\times$    & \citet{Lister:2011kg}       \\
ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp     & $26\times$    & \citet{Lister:2011kg}       \\
FF                  & Foreskin fibroblasts    & SE         & $85$ bp     & $16\times$    & \citet{Lister:2011kg}       \\
FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp     & $10\times$    & \citet{Lister:2011kg}       \\
FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp     & $8\times$     & \citet{Lister:2011kg}       \\
FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp     & $17\times$    & \citet{Lister:2011kg}       \\
IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
H1\_r1              & ESC                     & SE         & $85$ bp     & $15\times$    & \citet{Lister:2009hy}       \\
H1\_r2              & ESC                     & SE         & $85$ bp     & $14\times$    & \citet{Lister:2009hy}       \\
H1+BMP4             & IVD from H1             & SE         & $85$ bp     & $33\times$    & \citet{Lister:2011kg}       \\
H9                  & ESC                     & SE         & $85$ bp     & $9\times$     & \citet{Lister:2011kg}       \\
H9\_Laurent         & ESC                     & PE         & $75$ bp     & $8\times$     & \citet{Laurent:2010iw}      \\
HSF1                & ESC                     & SE         & $47$        & $5\times$     & \citet{Chodavarapu:2010iqa} \\ \bottomrule
\end{tabulary}
\end{table}

>>>>

The methylC-seq libraries from the \cite{Lister:2009hy} paper were the first published whole-genome bisulfite-sequencing libraries of mammalian DNA. A focus of this paper was comparing DNA methylation levels in a somatic tissue, fetal lung fibroblasts (_IMR90_), with those from a pluripotent tissue, embryonic stem cells (_H1_). Each tissue was run in duplicate. While \cite{Lister:2009hy} refer to these "biological" replicates I believe that these are better described as technical replicates since each replicate is from the same cell line; what distinguishes the replicates are the number of cell passages and the subsequent library preparations and sequencing. In any case, the published analyses pool these duplicates, which ignores all between-replicate variability. These samples are detailed in Table \ref{tab:Lister_dataset_pt1}.

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset part 1]{Samples in \emph{Lister} dataset from \cite{Lister:2009hy}. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{ESC} = embryonic stem cell; \emph{SE} = single-end sequencing.}
\label{tab:Lister_dataset_pt1}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
IMR90\_r1           & Lung fibroblasts        & SE         & $87$ bp    & $14\times$ \\
IMR90\_r2           & Lung fibroblasts        & SE         & $87$ bp    & $15\times$ \\
H1\_r1              & ESC                     & SE         & $85$ bp    & $15\times$ \\
H1\_r2              & ESC                     & SE         & $85$ bp    & $14\times$ \\
\bottomrule
\end{tabulary}
\end{table}

The methylC-seq libraries from the \cite{Lister:2011kg} include some created by the authors and some published by other groups. These samples include cell lines from differentiated cell lines, embryonic stem cell lines, pluripotent stem cell lines and _in vitro_ differentiated from pluripotent stem cells. There are no replicates for any of the \cite{Lister:2011kg} samples. These samples are detailed in Table \ref{tab:Lister_dataset_pt2}.

\begin{table}[!htb]
\centering
\caption[Description of \emph{Lister} dataset part 2]{Samples in \emph{Lister} dataset from \cite{Lister:2011kg}. The reported read lengths are prior to any trimming of the reads. Abbreviations: \emph{iPSC} = induced pluripotent stem cell; \emph{ESC} = embryonic stem cell; \emph{IVD} = \emph{in vitro} differentiated from pluripotent cell line; \emph{SE} = single-end sequencing; \emph{PE} = paired-end sequencing.}
\label{tab:Lister_dataset_pt2}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
Sample              & Tissue type             & Sequencing & Read length & Ave. coverage \\
ADS                 & Adipose                 & PE         & $75$ bp    & $23\times$ \\
ADS-adipose         & Adipocytes from ADS     & PE         & $75$ bp    & $24\times$ \\
ADS-iPSC            & iPSC from ADS           & PE         & $75$ bp    & $26\times$ \\
FF                  & Foreskin fibroblasts    & SE         & $85$ bp    & $16\times$ \\
FF-iPSC\_6.9        & iPSC from FF            & SE         & $85$ bp    & $10\times$ \\
FF-iPSC\_19.7       & iPSC from FF            & SE         & $85$ bp    & $9\times$  \\
FF-iPSC\_19.11      & iPSC from FF            & SE         & $85$ bp    & $8\times$  \\
FF-iPSC\_19.11+BMP4 & IVD from FF-iPSC\_19.11 & SE         & $85$ bp    & $17\times$ \\
IMR90-iPSC          & iPSC from IMR90         & SE         & $85$ bp    & $9\times$ \\
H1+BMP4             & IVD from H1             & SE         & $85$ bp    & $33\times$ \\
H9                  & ESC                     & SE         & $85$ bp    & $9\times$ \\
H9\_Laurent         & ESC                     & PE         & $75$ bp    & $8\times$ \\
HSF1                & ESC                     & SE         & $47$ bp    & $5\times$ \\
\bottomrule
\end{tabulary}
\end{table}

\FloatBarrier

### 2015-05-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Fix table formatting

\begin{table}[h]
\centering
\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data. All m-tuples are stranded. The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The `number of rows' corresponds to the number of m-tuples in the object. The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples. The final column is a measure of how sparse the data are: a $0$ values means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
& Size (GB) & Number of rows & Number of assays & Percentage of $0$ and \texttt{NA} values \\ \midrule
1-tuples                                 & $5.9$                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$                       & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{{-}{-}all-combinations}) & $60.0$                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$                       & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$                       & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
\end{tabulary}
\end{table}

>>>>

\begin{table}[h]
\centering
\caption[Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data]{Size of \texttt{MethPat} objects for the \emph{EPISCOPE} data. All m-tuples are stranded. The `size' of the \texttt{MethPat} object, reported in gigabytes (GB), is computed using the \texttt{pryr::object\_size()} function (\url{http://cran.r-project.org/web/packages/pryr/index.html}). The `number of rows' corresponds to the number of m-tuples in the object. The `number of assays' is $2^{\text{m}}$, where m is the size of the m-tuples. The final column is a measure of how sparse the data are: a $0$ value means that particular methylation pattern was not observed in that particular sample and an \texttt{NA} value means that that m-tuple was not observed in that particular sample.}
\label{tab:methpat_sparsity}
\begin{tabulary}{\textwidth}{LLLLL}
\toprule
& Size (GB) & Number of rows & Number of assays & Percentage of $0$ and \texttt{NA} values \\ \midrule
1-tuples                                 & $5.9$                        & $56,348,522$    & $2$              & $28\%$                                     \\
2-tuples                                 & $20.1$                       & $100,586,237$   & $4$              & $80\%$                                     \\
2-tuples (\texttt{{-}{-}all-combinations}) & $60.0$                          & $299,814,999$   & $4$              & $78\%$                                     \\
3-tuples                                 & $43.3$                       & $109,376,348$   & $8$              & $93\%$                                     \\
4-tuples                                 & $80.5$                       & $102,625,758$   & $16$             & $97\%$                                       \\ \bottomrule
\end{tabulary}
\end{table}
